The markings of the trade: exploring the patterns of olive oil production in Roman Baetica

### 2020-07-22 Paper/jaccardbaetica.tex
add new changes xavi

The creation of new provinces required the Roman Empire the arrival of resources through Mediterranean and Atlantic routes.
%\xavi{no entiendo esto...allowed o required?}. 
This led to a gradual change in the economic and social structures the trade networks that supplied the Roman army became more complex. Augustus's administration created the figure of the \textit{praefectura annonae} to organize wheat supply. The role of the \textit{praefectura annonae} was mainly focused on providing through \textit{frumentationes} a fixed monthly amount of wheat to each Roman citizen~\citep{remesal_annona_1986,remesal_concierto}\xavi{pero estabas hablando del ejército no? Has hecho un salto aquí a los ciudadanos}.

Some hypotheses argue elsewhere that this same \textit{Annona} could also have organized the supply of additional goods such as olive oil to the Roman legions \citep{remesal_annona_1986,remesal_annona_1990}

%\xavi{pero no hay evidencia? Las fuentes escritas no lo mencionan? - He cambiado authors por hipotesis porque me quería referir a investigadores zorry}.

@@

Roman studies exploring the presence of Dressel 20 amphorae in \textit{Germania} has not had the same impact as the rest of the province\xavi{no entiendo esta frase. Que qué provincia? Germania son 2 provincias no?}. This could be explained due to the lack of archaeological sources \citep{horacio2010llegada}. As a consequence, it is still unknown if trade agents participated on the distribution of Baetican olive oil in \textit{Germania} or the supply was exclusively focused on the Roman army garrisons\citep[156]{remesal_germn_2010}\xavi{quieres decir que aquí sólo encontramos dressel 20 en yacimientos militares porque no se han excavado muchos civiles?}.

@@

The presence of the Roman army encouraged the exchange in the province as shown by the arrival of this product both civil settlement and military sites with a mayor concentration at the German \textit{limes}\xavi{pero justo antes decías que no había en civiles no?}. It seems that some Baetican centres would be assigned to the support of olive oil\xavi{the support of olive oil no se entiende. Además, falta alguna cita y desarrollar esta idea}. However, this hypothesis is hard to assess given the current lack of archaeological evidence \xavi{related to BLABLABLA. Habría que detallar un poco porque hay un montón de evidencia arqueológica del limes, así que qué es exactamente lo que falta?}\citep[125]{remesal_concierto}.

@@

The first step was to compute the frequency distribution of stamp codes and analyse it using Exploratory Data Analysis (EDA). This would allow us to study the distribution of the stamps across centres as can be seen in Figure~\ref{stamps}. Most workshops contained only one stamp~\xavi{el gráfico es dificil de entender...yo leo que hay unos 20 workshops con 1 solo stamp, por lo que entiendo que no es cierto que la mayoría solo tengan 1 sello (hay como 50 workshops con entre 2 y 30 sellos no?)} One workshop (La Catria) concentrated a large percentage of the amphoric stamps with a total number of 228 stamps. The type of distribution is also frequent in amphora production where we observed a self-organised complex system pattern with a major concentration of the number of stamps in a few workshops \citep{bayesian_2018,coto-sarmiento_identifying_2018}\xavi{no lo veo; recuerda que para el rollo bayesiano tenías una línea recta al hacedr el log de los 2 ejes, no solo 1, y este plot no está construído así no?}.

%The type of distribution is also frequent in the Roman economy where we observed a self-organised complex system pattern typical of a free market

%\xavi{que me quede claro...la mayoría de workshops solo tienen 1 sello??? Y hay un workshop con como 700 sellos distintos??? }

@@

By contrast, most sites did not show a strong similarity in stamp correlation\xavi{qué quieres decir? porque lo de que hay una similaridad baja ya lo has repetido arriba}. We also did not find any groups of similar stamps in a specific place\xavi{tampoc entiendo esto. que no hay sellos similares en un solo sitio...?}.

These results suggest that there is no clear organizational pattern linking production centres and/or consumption centres beyond spatial distance. As a consequence, our analysis does not indicate the presence of a specific production centre from \textit{Baetica} to \textit{Britannia} province. Rather, it would imply that olive oil production was distributed by non-specific production centres\xavi{esto no iría en la discusión? Por otra parte, no entiendo qué quieres decir exactamente, igual hay que desarrollarlo}.

@@

A higher concentration of sites sharing similar stamps were found in areas eminently militarised and close to German \textit{limes}, even if most sites mostly showed different stamps~\xavi{quieres decir que los sitios más cercanos al limes son más similares? no queda claro y hay que detallarlo un poco más}.

@@

The analysis of the amphora workshops in \textit{Baetica} province suggests that a majority of stamp codes were only used on a single workshop. Beyond this clear pattern it seems that the similarity of stamp codes is correlated with spatial distance and the amphoric production of spatially close sites shared some stamp codes. Particularly relevant here is the fact that our results show how similar stamp codes were found on sites belonging to the same \textit{conventus} area. This result could indicate that the workshops of the different \textit{conventus} used different stamp codes that were not shared between them. Despite these stamps tend to share the same area of production, we do not identify groups of any more than three workshops sharing the same amphoric stamps as the dendrogram showed. In general, the majority of stamps were located in different amphora workshops\xavi{estas 2 frases repiten cosas dichas antes}.

Our results do not support previous working hypotheses suggesting that groups of workshopes used specific groups of stamp codes; while there is higher stamp similarity for nearby workshops this result does not suggest any kind of organisation beyond the simple fact that nearby workshops had a higher degree of interaction. In fact; the opposite seems true: each workshop used different stamp codes and only shared some of them with the closest ones. These stamp codes detected in closer workshops do not move from other distant workshops. In other words, the stamps tend to remain in the same area and different stamps were located in a same amphora workshop\xavi{de nuevo estas 2 frases repiten cosas de antes}.

This unique link between workshop and stamp code can be interpreted in different ways. First, it is possible that each workshop operated independently and did not normally collaborate with other workshops.  Second, stamp similarity in closer workshops could be linked to a spatial pattern. It is more probable than closer workshops tend to share more traits than distant workshops. While the role of the river was significant for the distribution of amphorae, river connectivity amongst workshops does not seem to show relevancy for the distribution of stamps\xavi{qué quieres decir con esto? que los diferentes ríos no muestran ningún patrón?}.

@@

This could be explained such as batch systematic organisation between potters \citep{juanmorostesis}. This method allowed to potters organises the production with the batch stamping for the posterior commercialisation\xavi{y esto para qué serviría? debería explicarse un poco más}.

@@

A combined exploration of the map and the dendrogram also suggests an interesting pattern: centres with higher similarity values are closer to the coast (North Sea and Celtic Sea). This may indicate that the Atlantic route could have played an essential role in transporting olive oil to the area of \textit{Britannia}, since in the places where there is greater similarity they are found in different strategic points near the sea\xavi{por qué más conectividad implica más similaridad? no será porque encuentras más ánforas?}.

@@

>>>>

The creation of new provinces required the Roman Empire the arrival of resources through Mediterranean and Atlantic routes.
%\xavi{no entiendo esto...allowed o required?}. 
This led to a gradual change in the economic and social structures the trade networks that supplied the Roman army became more complex. Augustus's administration created the figure of the \textit{praefectura annonae} to organize wheat supply. The role of the \textit{praefectura annonae} was mainly focused on providing through \textit{frumentationes} a fixed monthly amount of wheat to each Roman citizen~\citep{remesal_annona_1986,remesal_concierto}\xavi{pero estabas hablando del ejército no? Has hecho un salto aquí a los ciudadanos}. 
\maria{Pero porque la annona estaba enfocada a los ciudadanos romanos y luego hablo después que hay otras hipótesis que dicen que iba más allá a los militares}

Some hypotheses argue elsewhere that this same \textit{Annona} could also have organized the supply of additional goods such as olive oil to the Roman legions \citep{remesal_annona_1986,
remesal_annona_1990}

\xavi{pero no hay evidencia? Las fuentes escritas no lo mencionan?}
\maria{He cambiado authors por hipotesis porque me quería referir a investigadores}.

@@

Roman studies exploring the presence of Dressel 20 amphorae in the provinces Inferior and Superior have not had the same impact as the rest of the provinces.
%\xavi{no entiendo esta frase. Que qué provincia? Germania son 2 provincias no?}

This could be explained due to the lack of archaeological sources \citep{horacio2010llegada}. As a consequence, it is still unknown if trade agents participated on the distribution of Baetican olive oil in \textit{Germania} or the supply was exclusively focused on the Roman army garrisons\citep[156]{remesal_germn_2010}\xavi{quieres decir que aquí sólo encontramos dressel 20 en yacimientos militares porque no se han excavado muchos civiles?}.

@@

The presence of the Roman army encouraged the exchange in the province as shown by the arrival of this product both civil settlement and military sites with a mayor concentration at the German \textit{limes}\xavi{pero justo antes decías que no había en civiles no?}. It seems that some Baetican centres would be assigned to the support of olive oil\xavi{the support of olive oil no se entiende. Además, falta alguna cita y desarrollar esta idea}. However, this hypothesis is hard to assess given the current lack of archaeological evidence related to 
\xavi{related to BLABLABLA. Habría que detallar un poco porque hay un montón de evidencia arqueológica del limes, así que qué es exactamente lo que falta?}\citep[125]{remesal_concierto}.

@@

The first step was to compute the frequency distribution of stamp codes and analyse it using Exploratory Data Analysis (EDA). This would allow us to study the distribution of the stamps across centres as can be seen in Figure~\ref{stamps}. Most workshops concentrated few stamps while few workshops contained more than 100 stamps. For instance, one workshop (La Catria) concentrated a large percentage of the amphoric stamps with a total number of 228 stamps.

%\xavi{el gráfico es dificil de entender...yo leo que hay unos 20 workshops con 1 solo stamp, por lo que entiendo que no es cierto que la mayoría solo tengan 1 sello (hay como 50 workshops con entre 2 y 30 sellos no?)} 
\maria{he cambiado la frase a ver si se entiende mejor}

The type of distribution is also frequent in amphora production where we observed a self-organised complex system pattern with a major concentration of the number of stamps in a few workshops \citep{bayesian_2018,coto-sarmiento_identifying_2018}\xavi{no lo veo; recuerda que para el rollo bayesiano tenías una línea recta al hacedr el log de los 2 ejes, no solo 1, y este plot no está construído así no?}.
\maria{son dos ejes, no? ¿aquí no se ve línea recta? Lo mismo si lo hago de puntitos...}

@@

%By contrast, most sites did not show a strong similarity in stamp correlation\xavi{qué quieres decir? porque lo de que hay una similaridad baja ya lo has repetido arriba}. We also did not find any groups of similar stamps in a specific place\xavi{tampoc entiendo esto. que no hay sellos similares en un solo sitio...?}.

@@

A higher concentration of sites sharing similar stamps were found in areas eminently militarised and close to German \textit{limes}, although most sites mostly showed different stamps. 
\maria{he cambiado esta frase}
%even if most sites mostly showed different stamps~\xavi{quieres decir que los sitios más cercanos al limes son más similares? no queda claro y hay que detallarlo un poco más}.

@@

The analysis of the amphora workshops in \textit{Baetica} province suggests that a majority of stamp codes were only used on a single workshop. Beyond this clear pattern it seems that the similarity of stamp codes is correlated with spatial distance and the amphoric production of spatially close sites shared some stamp codes. Particularly relevant here is the fact that our results show how similar stamp codes were found on sites belonging to the same \textit{conventus} area. This result could indicate that the workshops of the different \textit{conventus} used different stamp codes that were not shared between them. Despite these stamps tend to share the same area of production, we do not identify groups of any more than three workshops sharing the same amphoric stamps as the dendrogram showed.

%In general, the majority of stamps were located in different amphora workshops\xavi{estas 2 frases repiten cosas dichas antes}. 
\maria{he quitado la última frase pero la anterior no se repite}

Our results do not support previous working hypotheses suggesting that groups of workshops used specific groups of stamp codes; while there is higher stamp similarity for nearby workshops this result does not suggest any kind of organisation beyond the simple fact that nearby workshops had a higher degree of interaction. In fact; the opposite seems true: each workshop used different stamp codes and only shared some of them with the closest ones.

%These stamp codes detected in closer workshops do not move from other distant workshops. In other words, the stamps tend to remain in the same area and different stamps were located in a same amphora workshop\xavi{de nuevo estas 2 frases repiten cosas de antes}.

This unique link between workshop and stamp code can be interpreted in different ways. First, it is possible that each workshop operated independently and did not normally collaborate with other workshops.  Second, stamp similarity in closer workshops could be linked to a spatial pattern. It is more probable than closer workshops tend to share more traits than distant workshops. While the role of the river was significant for the distribution of amphorae, river connectivity amongst workshops does not seem to show relevancy for the distribution or concentration of stamps in a specific river area.
\xavi{qué quieres decir con esto? que los diferentes ríos no muestran ningún patrón?}
\maria{he añadido cosas a la frase para ver si se entiende}

@@

This could be explained such as batch systematic organisation between potters \citep{juanmorostesis}. This work method could have allowed to potters to stamp amphorae in order to identify the production for batches for posterior commercialization. High production of amphorae had to maintain a specialize system organization to control this production \citep[104]{juanmorostesis}.

%This method allowed to potters organises the production with the batch stamping for the posterior commercialisation
%\xavi{y esto para qué serviría? debería explicarse un poco más}.

@@

A combined exploration of the map and the dendrogram also suggests an interesting pattern: centres with higher similarity values are closer to the coast (North Sea and Celtic Sea). This may indicate that the Atlantic route could have played an essential role in transporting olive oil to the area of \textit{Britannia}, since in the places where there is greater similarity they are found in different strategic points near the sea
\xavi{por qué más conectividad implica más similaridad? no será porque encuentras más ánforas?}
\maria{no entiendo este comentario :-(}.

@@

Finally, the results suggest that there is no clear organizational pattern linking production centres and/or consumption centres beyond the spatial distance. As a consequence, our analysis does not indicate the presence of a specific workshop from \textit{Baetica} supplying to some specific province.

\xavi{esto no iría en la discusión? Por otra parte, no entiendo qué quieres decir exactamente, igual hay que desarrollarlo}.
\maria{lo he incluido aquí y lo he cambiado a ver si se entiende}
%Rather, it would imply that olive oil production was distributed from non-specific production centres.
